---
title: "TA Session 3b"
author: "Jena Manilla"
date: "April 23, 2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# This document is intended for educational purposes only for
# Fiona Burlig's PPHA 34600 Program Evaluation Course at the
# Harris School of Public Policy

# TA Session 3b details the following:
#
# 1. Intro to RMarkdown
# 2. Data Wrangling
# 3. Regression 101
# 4. Tables and Graphs
#
library(knitr)
library(tidyverse)
library(dplyr)

knitr::opts_chunk$set(echo = TRUE)
```

## 1. R Markdown

If you've opened this file in R, well done! You are well on your way to using R Markdown, as this is an R Markdown document. If you had trouble opening this document (or you opened it as a PDF), make sure you have R Studio installed.  

### To install R Studio  

1. Go to www.rstudio.com and click on the "Download" button on the top right (above "Pricing".)  
2. Click on the "Download R Studio Desktop"  
3. Click on the version recommended for your system, or the latest Windows version, and save the executable file. Run the .exe file and follow the installation instructions.  

RMarkdown is part of R Studio. To open a new R Markdown document, select "File" --> "New File" --> "R Markdown". Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

You can embed an R code chunk like this:

```{r, warning=FALSE, message=FALSE}
# You can type code or text here! This gray chunk works just like a normal .R file
# Let's read in some data
burrito <- read_csv('https://raw.githubusercontent.com/BUSN32100/data_files/master/burrito.csv')
```
This is the advantage of R Markdown. The gray chunk above allows you to code, while this white space works for free text. There are lots of ways to format in R Markdown, including *italics* and **bold**. You can also type equations, either in-line like this $y = mx + b$ or on their own line
$$ y = mx + b $$

Once you're done with your R Markdown file, click the **Knit** button and a document will be generated that includes both text content as well as the output of any embedded R code chunks within the document. 
 
*Big caveat: there is no one right way code! These are only suggestions.*  

## 2. Data Wrangling  
A lot of this class requires filtering and subsetting dataframes. One helpful tool for this is **pipes**, which looks like this %>%     

When is the pipe operator helpful? Here's an example. Suppose I want to:  

 1. find keys, 
 2. start car, 
 3. drive to school, 
 4. park.  

I can write this as one line:
```{r eval=FALSE}
park(drive(start_car(find("keys")), to = "campus"))
```

But writing it out using pipes makes it easier to read:
```{r eval=FALSE}
find("keys") %>%
  start_car() %>%
  drive(to = "campus") %>%
  park()
```

The pipes operator is also very helpful for maniplating data sets. Here are some examples of what you can do:  
```{r, eval=FALSE}
# Filter based on criteria
burrito %>% 
  filter( Burrito == "carnitas", overall >4) #filter for burritos with 
                                                #carnitas and an overall rating > 4

# Select specific columns.
burrito %>% 
  select(Location, Burrito, Meat, Salsa) # Return 4 columns using their name

burrito %>% 
  select(Location:Salsa) # Return all columns between location and salsa

# Drop columns by adding a minus sign in front of the column name.
burrito %>% 
  select(-Location) # This gets rid of location

# Make new columns with the mutate function
burrito %>% 
  select(Burrito, overall) %>% # Select two columns
  mutate(overall_double = overall*2) %>% # Create a new column called overall_double
  head() # Show the first few rows of the new dataframe

```
Those examples are really just the tip of the iceberg for the pipes operator! You can find a lot of examples via stack overflow and Google.    

Also, here's a **pro tip**: use `ctrl/cmd + shift + m` as a shortcut.  

## 3. Regression 101  

There are lots of ways to write code for regressions. Here is just one example: 
```{r, comment=NA, eval=FALSE}
regH <- lm(overall ~ Hunger, burrito)
summary(regH)
coef(summary(regH)) # Isolate just the coefficients from the regression
coef(summary(regH))["Hunger", c("Estimate", "Pr(>|t|)")] # Isolate treatment effect

```

Here is an example of how to run many regressions at once: 
```{r, comment=NA, eval=FALSE}
# Run regression of overall on tortilla, meat and filling
regs <- lapply(burrito %>% 
                 select(Tortilla, Meat, Fillings), function(x) lm(overall ~ x, burrito))
# Let's see what the output is
regs
# Call the coefficients on Tortilla
coef(summary(regs$Tortilla))["x", c("Estimate", "Pr(>|t|)")]
# Call all the coefficients
t(sapply(regs, function(x) coef(summary(x))["x", c("Estimate", "Pr(>|t|)")]))

```



## 4. Tables and Graphs

It's often a good idea to visualize your data. A really great package for this is ggplot.
```{r}
# Make sure you install and/or load ggplot
library(ggplot2)

```

Let's plot the distribution of the overall burrito ratings:
```{r, fig.width=4,fig.height=4}
ggplot(data = burrito, mapping = aes(overall)) + geom_histogram(binwidth = .1)

```

And now a scatterplot of fillings vs overall rating, with a line of best fit:
```{r, warning=FALSE, comment=FALSE}
ggplot(data = burrito, mapping = aes(x=Fillings, y=overall)) + 
  geom_point() + geom_smooth(method = lm, se=F)
```

These are just two simple examples. Play around with colors, titles, different types of plots... there are lots of options!  
  
\newpage  

Stargazer is also a very helpful tool for making nice tables. https://cran.r-project.org/web/packages/stargazer/vignettes/stargazer.pdf  

For example, here's a table with our burrito regressions from before:
```{r, warning=FALSE, message=FALSE, comment=NA}
regs <- lapply(burrito %>% 
                 select(Tortilla, Meat, Fillings), function(x) lm(overall ~ x, burrito))
library(stargazer)
stargazer(regs, type = "text", column.labels = c("Tortilla", "Meat", "Filling"))

```

