---
title: 'TA Session #9'
author: "null"
date: "June, 2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(plm)
library(haven)
library(readr)
library(stargazer)
library(MatchIt)
library(AER)
library(lfe)
library(plm)
```

## Randomized Control Trials (RCT)

#### Checking for balance (via Regression)

Let's use the data from PS1. We want to check for balance in *baseline wages, baseline sales, baseline employees, and baseline owner (female).* The treatment variable is *burlig_trt*

```{r}
# Read data
rm(list=ls())
ps1_data <- read_csv("ps1_data.csv")

# Run regression of covariates on treatment to check for balance
balance_wages <- lm(baseline_wages ~ burlig_trt, data = ps1_data)
balance_sales <- lm(baseline_sales ~ burlig_trt, data = ps1_data)
balance_employees <- lm(baseline_employees ~ burlig_trt, data = ps1_data)
balance_owner_female <- lm(baseline_owner_female ~ burlig_trt, data = ps1_data)

#Use stargazer to summarize regression outputs
stargazer(balance_wages, balance_sales, balance_employees, balance_owner_female, type = "text")
```


## 2. Matching

#### 2.1 Exact Matching

This matching method shows how to split the dataset into cells by taking unique combinations of your choice of covariates.

```{r}
# Read data
rm(list=ls())
ps2_data <- read_csv("ps2_data.csv")

# Clean data
ps2_data$farmer_birth_year[which(ps2_data$farmer_birth_year=="nineteen seventy-two")] <- 1972
ps2_data$farmer_birth_year[which(ps2_data$farmer_birth_year=="nineteen seventy-three")] <- 1973

# (1) create a "weight" column. weight = 1 for each observation 
# (2) group by variables to match
df_ps2 <- ps2_data %>%
  mutate(weight = 1) %>%
  group_by(fiona_farmer, farmer_birth_year) %>%
  summarise(profits = mean(profits_2016),
            weight = sum(weight))

# Check for Treatment and Control matches
treated <- df_ps2 %>%
  filter(fiona_farmer == 1)

untreated <- df_ps2 %>%
  filter(fiona_farmer == 0)

# Filter out unmatched units
df_ps2 <- df_ps2 %>%
  group_by(farmer_birth_year) %>%
  filter(n()>1)

# check again
treated <- df_ps2 %>%
  filter(fiona_farmer == 1)

untreated <- df_ps2 %>%
  filter(fiona_farmer == 0)

# Estimate ATE

data <- cbind(treated, untreated)

# Difference in treatment effect
data$TE <- data$profits - data$profits1
  
# For how many observations?
data$weight_total <- data$weight + data$weight1

# Apply Matching's ATE Formula

data$ATE <- data$TE * (data$weight_total / sum(data$weight_total))

# What is the ATE?
ATE <- sum(data$ATE)
ATE

```

## 3. Instrumental Variables (IV)

#### 3.1 2SLS

Be careful that the standard error in 2SLS is *not correct*. We will get the correct standard error in the next method.

```{r first_stage}
# Read data
rm(list=ls())
card <- read_csv("card.csv")

# First stage
first <- lm(educ ~ nearc4 + exper + expersq, data = card)
summary(first)

# Save predicted values
card$fit_first <- fitted.values(first)
```

```{r second_stage}
# Second Stage
second <- lm(lwage ~ fit_first + exper + expersq, data = card)
summary(second)
```

IV estimate can also be calculated using first stage regression and reduced form:

```{r reduced_form}
# Reduced form
theta <- coef(lm(lwage ~ nearc4 + exper + expersq, data = card))
gamma <- coef(lm(educ ~ nearc4 + exper + expersq, data = card))
tau <- theta / gamma
tau

```

#### 3.2 ivreg

There is a package in R *"AER"*. You can use the *ivreg* function to get the IV estimate with the correct standard error. 

```{r}

iv <- ivreg(lwage ~ educ + exper + expersq |
  nearc4 + exper + expersq, # don't forget to also put covariates here
data = card
)
summary(iv)
```


## 4. Difference in Difference (DiD)


```{r}
rm(list=ls())
evdata <- read.csv("evdata_sample.csv")

# DiD by simply comparing means

# Treated units before and after
trt_after<- evdata %>%
  filter(evdata$treat == 1 & evdata$post == 1)

trt_before <- evdata %>%
  filter(evdata$treat == 1 & evdata$post == 0)
                       
# Control units before and after
ctrl_after<- evdata %>%
  filter(evdata$treat == 0 & evdata$post == 1)

ctrl_before <- evdata %>%
  filter(evdata$treat == 0 & evdata$post == 0)

# Now, first difference for treated and control
Treated_mean <- mean(trt_after$electricity_use) - mean(trt_before$electricity_use)
control_mean <- mean(ctrl_after$electricity_use) - mean(ctrl_before$electricity_use)

# Now the Difference in Difference
DID <- Treated_mean - control_mean
DID

###############################################################################

# DiD using regression (no fixed effects)?coeftest
DiD_Reg <- lm(electricity_use ~ treat + post + treat*post, data = evdata)
summary(DiD_Reg)

```

The coefficient for the *interaction term* will be our DiD estimate.

## 5. Fixed Effects

```{r}
# Generate the treatment status variable
evdata$D <- ifelse(is.na(evdata$ev_purchase_year) | (evdata$year<2013), 0, 1)

# Fixed effects regression using plm
fe <- plm(electricity_use ~ D,
          data = evdata,
          model = "within", # setting for de-meaining within individual fixed effect model
          index = c('hh_id', 'year'), # specify the individual and time index
          effect = "twoways") 
summary(fe)
coeftest(fe, vcov=vcovHC(fe, type="HC0", cluster="group")) # Get the clustered std.Error
```

